Computing likelihoods under the SMC for a general class of ARGs.

### 2024-07-05 paper.tex
More updates on the intro

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Applications using inferred ARGs as input have begun to
appear~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024}.
Additionally, there is a rich suite of tools growing around
the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
adrion_community_2020,terasaki_geonomics_2021,
baumdicker_efficient_2021,
korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
wohns_unified_2022,mahmoudi_bayesian_2022,
rasmussen_espalier_2022,
zhang_biobank-scale_2023,zhan_towards_2023} and analysis
packages
% etc
\citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
Infererence methods now tend to fall into two classes: those that
sample from a statistical
model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022,
deng_robust_2024}
or parsimony~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022},
and those based on
heuristics~\citep{kelleher_inferring_2019,speidel_method_2019,
zhang_biobank-scale_2023}.
Typically, the more rigorous approaches can handle 10-100 samples,
while the approximate methods can scale to 10s to hundreds of
thousands of samples. With the explosive growth in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference.

Perhaps the clearest route to statistical rigor is through scalable
model-based inference, a central element of population genetics.
The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
tajima_evolutionary_1983} has proved to be a useful model, and is the
basis of a wide variety of inferences [some citations, e.g Hey IM model]. 
The incorporation
of recombination in the coalescent process~\citep{hudson_properties_1983}
has proved to be much more difficult.
% FIXME rephrase, next two sentences, lifted from ARG paper
Early work on inference under the coalescent with recombination
focused on the problem of
inferring the parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths_ancestral_1996,kuhner_maximum_2000,
nielsen_estimation_2000,
fearnhead_estimating_2001}.
% TODO reuse some of the ideas from the SMC paper here, it has
% some good explanations.
These methods met with limited success
because the state space of ARGs is overwhelmingly large, and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The introduction of the Sequentially Markovian Coalescent (SMC)
approximation to the coalescent with recombination
by~\cite{mcvean_approximating_2005} was a seminal contribution,
and has made many different inferences
% TODO add some more recent stuff
feasible~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
In particular, the SMC is the basis for ARG sampling methods
like ARGweaver~\citep{rasmussen_genome-wide_2014}, which remains the
% TODO list other places ARGweaver has come out on top in terms of accuracy.
gold-standard in terms of accuracy~\citep{brandt2022evaluation}

% THE PROBLEM!!! OMG it's terrible!!!
Although major breakthroughs in scale have been achieved by hueristic
ARG inference
methods~\citep{kelleher_inferring_2019,speidel_method_2019,zhang_biobank-scale_2023},
these methods are currently fundamentally incompatible with rigorous
inference under population genetic models such as the SMC. This is 
because these methods return approximate ARG
\emph{structures}~\citep{wong_general_2023} [WEAK, FIXME]. Specifically,
we cannot evaluate a likelihood of these ARGs in coalescent 
models under the classical 
\cite{kuhner_maximum_2000} formulation, as these formulations assume
that complete information about recombination events is present.
[NOTE the following is super weak, and needs a lot of work.]
This inability to connect the results of large-scale inference methods
with a generative model
is a fundamental limitation, and limits both the utility of 
current point estimates as well as the possibilities for further
inference. Specifically, we are unable to quantify uncertainty 
around heuristic point estimates. We cannot explore ARG space
around the point estimates using (e.g. MCMC), as we have no way definition
of improvement. We cannot compare models of demography, or evaluate the 
goodness of fit for model estimates.

>>>>

\section{Introduction}
% TODO need to rewrite this as its too tskit-centric. We don't need to 
% emphasise tskit/msprime so much here, it's not necessary and will 
% annoy people. 
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Applications using inferred ARGs as input have begun to
appear~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024}.
Additionally, there is a rich suite of tools growing around
the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
adrion_community_2020,terasaki_geonomics_2021,
baumdicker_efficient_2021,
korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
wohns_unified_2022,mahmoudi_bayesian_2022,
rasmussen_espalier_2022,
zhang_biobank-scale_2023,zhan_towards_2023} and analysis
packages
% etc
\citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
Infererence methods now tend to fall into two classes: those that
sample from a statistical
model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022,
deng_robust_2024}
or parsimony~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022},
and those based on
heuristics~\citep{kelleher_inferring_2019,speidel_method_2019,
zhang_biobank-scale_2023}.
Typically, the more rigorous approaches can handle 10-100 samples,
while the approximate methods can scale to 10s to hundreds of
thousands of samples. With the explosive growth in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference.

Perhaps the clearest route to statistical rigor is through scalable
model-based inference, a central element of population genetics.
The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
tajima_evolutionary_1983} has proved to be a useful model, and is the
basis of a wide variety of inferences [some citations, e.g Hey IM model]. 
The incorporation
of recombination in the coalescent process~\citep{hudson_properties_1983}
has proved to be much more difficult.
% FIXME rephrase, next two sentences, lifted from ARG paper
Early work on inference under the coalescent with recombination
focused on the problem of
inferring the parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths_ancestral_1996,kuhner_maximum_2000,
nielsen_estimation_2000,
fearnhead_estimating_2001}.
% TODO reuse some of the ideas from the SMC paper here, it has
% some good explanations.
These methods met with limited success
because the state space of ARGs is overwhelmingly large, and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The introduction of the Sequentially Markovian Coalescent (SMC)
approximation to the coalescent with recombination
\citep{mcvean_approximating_2005} 
and its refinement in the SMC'~\citep{marjoram_fast_2006},
are the basis of many powerful inference 
% TODO add some more recent stuff
methods~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
In particular, the SMC is the basis for ARG sampling methods
like ARGweaver~\citep{rasmussen_genome-wide_2014} and 
SINGER~\citep{deng_robust_2024}, which are the gold-standard 
in terms of inference accuracy.
ARGweaver has consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and, while SINGER has not been evaluated as thoroughly yet, it promises
to be even more accurate as well as substantially more scalable 
computationally~\citep{deng_robust_2024}.

% THE PROBLEM!!! OMG it's terrible!!!
Despite these recent advances in statistically rigorous ARG 
inference, however, there is still a gap of some three orders 
of magnitude between the sample-sizes that can be handled 
(hundreds) compared to the large-scale hueristic methods
like tsinfer and ARGneedle (hundreds of thousands).
[TODO working through a more positive narrative here, based on
the idea of combining the strengths of the approaches]
Although major breakthroughs in scale have been achieved by hueristic
ARG inference
methods~\citep{kelleher_inferring_2019,speidel_method_2019,zhang_biobank-scale_2023},
these methods are currently fundamentally incompatible with rigorous
inference under population genetic models such as the SMC. This is 
because these methods return approximate ARG
\emph{structures}~\citep{wong_general_2023} [WEAK, FIXME]. Specifically,
we cannot evaluate a likelihood of these ARGs in coalescent 
models under the classical 
\cite{kuhner_maximum_2000} formulation, as these formulations assume
that complete information about recombination events is present.
[NOTE the following is super weak, and needs a lot of work.]
This inability to connect the results of large-scale inference methods
with a generative model
is a fundamental limitation, and limits both the utility of 
current point estimates as well as the possibilities for further
inference. Specifically, we are unable to quantify uncertainty 
around heuristic point estimates. We cannot explore ARG space
around the point estimates using (e.g. MCMC), as we have no way definition
of improvement. We cannot compare models of demography, or evaluate the 
goodness of fit for model estimates.
